---
title: "Tvyal Newsletter"
date: "2024-08-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(rvest)
library(RcppRoll)
library(scales)
library(readxl)

# rm(list = ls()); gc()

setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

source("../../initial_setup.R")

```


```{r}
GDP_quarter <- read_csv("GDP_quarter_tidy.csv")

GDP_annual <- read_csv("GDP_annual_tidy.csv")

GDP_annual |> view()
```


```{r}
GDP_annual |> 
  filter(
    eng == "Domestic product (gross, market prices)",
    !is.na(vol_YoY_pct)
  ) |> 
  mutate(
    mean_groth = exp(mean(log(vol_YoY_pct), na.rm = TRUE)),
    across(c(mean_groth, vol_YoY_pct), ~.x/100 - 1),
    pct_text = percent(vol_YoY_pct, accuracy = 0.1)
  ) |> 
  ggplot(aes(year, vol_YoY_pct, label = pct_text)) +
  geom_col(fill = new_palette_colors[2], alpha = 1) +
  geom_hline(aes(yintercept = mean_groth), color = new_palette_colors[2], alpha = 0.6) +
  geom_hline(aes(yintercept = 0), color = new_palette_colors[1]) +
  geom_text(aes(
    vjust = ifelse(vol_YoY_pct >= 0.03, 1.5, -0.5),
    color = ifelse(abs(vol_YoY_pct) <= 0.03, "white", "blue"),
  )) +
  geom_text(
    data = tibble(x = 2013.5, y = 0.048, label = "4.4 ีดีซีปีซีถ ีกีณ"),
    aes(x, y, label = label)
  ) +
  scale_x_continuous(breaks = 2013:2024) +
  scale_y_continuous(n.breaks = 10) +
  scale_color_manual(values = c("white", new_palette_colors[2])) +
  labs(
    x = NULL,
    y = NULL,
    title = "ีีถีฟีฅีฝีกีฏีกีถ ีกีณีจ ีีกีตีกีฝีฟีกีถีธึีด",
    subtitle = "ีีซีปีซีถ ีฟีถีฟีฅีฝีกีฏีกีถ ีกีณีจ ีีกีตีกีฝีฟีกีถีธึีด 4.4% ีง",
    caption = caption_f("Armstat")
  ) +
  theme(
    legend.position = "drop",
    panel.grid.major.x = element_blank(),
    axis.text.y = element_blank(),
  )

```



```{r, include=FALSE}

arm_inflation <-
  read_csv("~/R/Gcapatker/2024_03_01_Inflation_forecst/inflation_forecast.csv") |>
  filter(
    indicator == "CPI",
    grepl("Actual|ets", model_name),
    period == "month"
  ) |>
  transmute(
    date = ym(paste(year(date), month(date))),
    # inflation = cumprod(value),
    inflation = roll_prodr(value, 12),
    inflation_model_name = model_name
  ) |>
  na.omit() |>
  arrange(date) |>
  group_by(date) |>
  slice_head(n = 1) |>
  ungroup()



GDP_quarter |>
  group_by(eng) |>
  mutate(
    vol_YoY_pct = vol_YoY_pct/100,
    vol_YoY_pct2 = production/lag(production, 4),
    # diff = vol_YoY_pct2 - vol_YoY_pct + 1,
    diff_2 = vol_YoY_pct2 / vol_YoY_pct,
    date = yq(date)
  ) |>
  ungroup() |>
  left_join(arm_inflation, by = "date") |>
  arrange(date) |>
  ggplot() +
  # geom_boxplot(aes(date, diff, group = date, color = "blue"), alpha = 0.4) +
  geom_boxplot(aes(date, diff_2, group = date, color = "red"), alpha = 0.4) +
  geom_line(aes(date, inflation))



# arm_inflation |> view()

GDP_annual <-
  read_csv("future_gdp_forecast_2.csv") |>
  group_by(eng) |>
  arrange(eng) |>
  mutate(
    year = year(date),
    month = month(date)
  ) |>
  ungroup() |>
  # count(date) |>
  left_join(arm_inflation, by = "date") |>
  mutate(.value_inf = .value / inflation) |>
  group_by(eng, year) |>
  summarise(
    production = sum(.value),
    production_inf = sum(.value_inf)
  ) |>
  ungroup() |>
  group_by(eng) |>
  mutate(
    vol_YoY_pct = production_inf / lag(production) * 100,
  ) |>
  ungroup() |>
  left_join(
    select(GDP_annual, code, eng) |> unique(), by = "eng"
  ) |>
  filter(year >= 2015) |>
  relocate(code)

# GDP_annual |> view()

GDP_annual |>
  filter(grepl("Domestic product", eng)) |>
  transmute(year, GDP_growth = vol_YoY_pct/100 - 1) |>
  na.omit()

read_csv("future_gdp_forecast.csv") |>
  filter(eng == "Domestic product (gross, market prices)") |>
  select(-contains("volume")) |>
  left_join(arm_inflation, by = "date") |>
  mutate(
    gdp_growth = .value,
    across(matches("value|conf"), ~.x/ inflation),
    year = year(date)
  ) |>
  group_by(year, .key) |>
  summarise(
    across(matches("value|conf|gdp"), ~sum(.x))
  )

```


```{r, include=FALSE}

read_csv("future_gdp_forecast.csv") |>
  group_by(eng) |>
  arrange(eng) |>
  mutate(
    vol_YoY_pct = .value / lag(.value, 4),
    year = year(date),
    month = month(date)
  ) |>
  filter(eng != "Value added (gross, at basic prices)") |>
  mutate(
    total = ifelse(eng == "Domestic product (gross, market prices)", TRUE, FALSE)
  ) |>
  group_by(year, total) |>
  summarise(.value = sum(.value)) |>
  arrange(total) |>
  group_by(total) |>
  mutate(growth = .value / lag(.value))

```




```{r constructing the main GDP plot, include=FALSE}

armenian_short_names <- tibble::tribble(
  ~code, ~arm_short,
  # NA, "ีีฅึึีซีถ ีกึีคีตีธึีถึ",
  # NA, "ิฑึีฟีกีคึีกีถึีซ ีฐีกึีฏีฅึ",
  # NA, "ิฑีพีฅีฌีกึีพีกีฎ ีกึีชีฅึ",
  # NA, "ีีิฑีิพ",
  "A", "ิณีตีธึีฒีกีฟีถีฟีฅีฝีธึีฉีตีธึีถ",
  "B", "ีีกีถึีกีฃีธึีฎีธึีฉีตีธึีถ",
  "C", "ิฑึีคีตีธึีถีกีขีฅึีธึีฉีตีธึีถ",
  "D", "ิทีฌีฅีฏีฟึีกีดีกีฟีกีฏีกึีกึีธึีด",
  "E", "ีึีกีดีกีฟีกีฏีกึีกึีธึีด",
  "F", "ีีซีถีกึีกึีธึีฉีตีธึีถ",
  "G", "ิฑีผึีฟีธึึ",
  "H", "ีีธีญีกีคึีธึีดีถีฅึ",
  "I", "ีีกีถึีกีตีซีถ ีฝีถีธึีถีค",
  "J", "ีีฅีฒีฅีฏีกีฟีพีธึีฉีตีธึีถ ึ ีฏีกีบ",
  "K", "ีีซีถีกีถีฝีถีฅึ",
  "L", "ิฑีถีทีกึีช ีฃีธึีตึ",
  "M", "ีีกีฝีถีกีฃีซีฟีกีฏีกีถ ีฃีธึีฎีธึีถีฅีธึีฉีตีธึีถ",
  "N", "ีีกึีนีกึีกึีธึีฉีตีธึีถ",
  "O", "ีีฅีฟีกีฏีกีถ ีฏีกีผีกีพีกึีธึีด",
  "P", "ิฟึีฉีธึีฉีตีธึีถ",
  "Q", "ิฑีผีธีฒีปีกีบีกีฐีธึีฉีตีธึีถ",
  "R", "ีีทีกีฏีธึีตีฉ ึ ีฐีกีถีฃีซีฝีฟ",
  "S", "ิฑีตีฌ ีฎีกีผีกีตีธึีฉีตีธึีถีถีฅึ",
  "T", "ีีถีกีตีซีถ ีฟีถีฟีฅีฝีธึีฉีตีธึีถีถีฅึ"
)
```

```{r, include=FALSE}

main_GDP_plot_data <- 
  GDP_quarter  |> 
  filter(!is.na(code)) |> 
  left_join(armenian_short_names, by = "code") |> 
  mutate(
    date = yq(date) + months(3) - days(1),
    year = year(date),
    eng = arm_short,
    eng = str_trunc(eng, 20),
    eng = paste0(code, ". ", eng),
    eng = fct_lump(eng, n = 7, w = production)
  ) |> 
  filter(year >= 2017) |> 
  group_by(date, year, eng) |> 
  summarise(production = sum(production), .groups = "drop") |> 
  mutate(code = str_replace_all(eng, "([A-Z])\\..*", "\\1")) |> 
  group_by(eng) |> 
  mutate(production_yoy = roll_sumr(production, 4)) |> 
  ungroup() |> 
  group_by(date) |>
  na.omit() |>
  mutate(
    pct = production_yoy/ sum(production_yoy),
  ) |> 
  ungroup() |>
  group_by(year) |> 
  mutate(
    pct_text = ifelse(
      eng != "Other" & date == max(date), # & year %in% c(2013, 2023), 
      paste0(
        number(production_yoy/1e6, accuracy = 0.01), "ึ (",
        percent(pct, accuracy = 0.1), ")", "โ"
        ),
      NA
    )
  ) |> 
  ungroup() |> 
  mutate(
    pct_text = case_when(
      date == max(date) & eng != "Other" ~ paste0("โ", eng,  " ", pct_text),
      date == min(date) & eng != "Other" ~ paste0(pct_text, code, "โ"),
      TRUE ~ pct_text
    ),
    production_yoy = production_yoy / 1e6
  )

factoring_levels <- 
  main_GDP_plot_data |> 
  filter(date == max(date)) |> 
  mutate(
    eng = fct_reorder(eng, production_yoy),
    eng = fct_relevel(eng, "Other", after = 0),
    eng = fct_rev(eng)
  ) |> 
  count(eng) |> 
  pull(eng)

main_GDP_plot_data <- 
  main_GDP_plot_data %>%
  mutate(eng = factor(eng, levels = factoring_levels))
         

segments <- 
  main_GDP_plot_data |> 
  # filter(year %in% c(2013, 2017, 2020, 2023)) |> 
  group_by(year) |> 
  filter(date == max(date)) |> 
  ungroup() |> 
  group_by(date) |> 
  summarise(
    production_yoy = sum(production_yoy),
    production_yoy_text = number(production_yoy, accuracy = 0.01),
    production_yoy = production_yoy + 1
  ) |>  
  ungroup() |> 
  mutate(
    production_yoy_text = ifelse(
      date == max(date), 
      paste0(production_yoy_text, "*"), 
      production_yoy_text
    ),
  )


# GDP_main_plot <- 
main_GDP_plot_data |> 
  ggplot(aes(date, production_yoy)) +
  geom_segment(
    data = segments,
    aes(x = date, y = production_yoy - 1, xend = date, yend = production_yoy),color="black"
  ) +
  geom_point(
    data = segments,
    aes(x = date, y = production_yoy),color="black"
  ) +
  geom_text(
    data = segments,
    aes(x = date, y = production_yoy + 0.5, label = production_yoy_text)
  ) +
  geom_area(aes(fill = eng)) +
  geom_text(
    data = main_GDP_plot_data |> filter(date == min(date)),
    aes(label = pct_text, color = eng),
    position = position_stack(vjust = 0.5), hjust = 1
  ) +
  geom_text(
    data = main_GDP_plot_data |> filter(date == max(date)),
    aes(label = pct_text, color = eng),
    position = position_stack(vjust = 0.5), hjust = -0
  ) +
  scale_x_date(
    date_breaks = "1 year", date_labels = "%Y", 
    # limits = c(NA_Date_, ymd("2025-01-01"))
  ) +
  scale_y_continuous(breaks = seq(0, 14, 2), labels = number_format()) +
  scale_fill_manual(values = new_palette_colors) +
  scale_color_manual(values = new_palette_colors) +
  coord_cartesian(clip = "off") +
  ggthemes::theme_fivethirtyeight() +
  labs(
    x = NULL,
    y = NULL,
    fill = NULL,
    title = "ิฑีถีฟีฅีฝีพีกีฎ ีฃีตีธึีฒีกีฟีถีฟีฅีฝีธึีฉีตีธึีถี 1-ีซีถ ีฟีฅีฒีซึ 4-ึีค",
    subtitle = "ีีิฑ-ีซ ีกึีฟีกีคึีธึีฉีตีธึีถีถ ีจีฝีฟ ีิณีิด ีญีดีข.2 ีคีกีฝีกีฏีกึีฃีดีกีถ, ีฟึีซีฌีซีธีถ ีคึีกีด",
    captions = caption_f("Armstat", suffix_text = "* 2024 ีฉีพีกีฏีกีถีซ ีกีผีกีปีซีถ ีฏีซีฝีกีดีตีกีฏีซ ีคึีธึีฉีตีกีดีข")
  ) +
  theme_minimal() +
  theme(
    axis.line.x = element_line(linewidth = .75),
    plot.title = element_text(size = 16),
    plot.subtitle = element_text(size = 12),
    panel.grid = element_blank(),
    axis.text.y = element_blank(),
    legend.position = "none",
    plot.margin = margin(10,220,10,110),
  )

```

in 2018 agriculture was the top GDP produser with 0.83 T drams, having 16.4% share, right now in 2024 it's in 4rd place, with 0.81 T drams, having 9% share. So it is less then in 2018, total degradation of agriculture and it went from the 1st place to 4th




```{r GDP change plot, include=FALSE}

# GDP_change_plot <-
GDP_quarter  |>
  filter(!is.na(code)) |>
  mutate(
    date = yq(date) + months(3) - days(1),
    year = year(date),
    eng = str_trunc(eng, 40),
    eng = paste0(code, ". ", eng),
  ) |>
  group_by(code, eng) |>
  mutate(production_yoy = roll_sumr(production, 4) / 1e3) |>
  filter(!is.na(production_yoy)) |>
  # filter(date %in% c(max(date), max(date) - years(1))) |>
  filter(date %in% c(max(date), min(date))) |>
  # filter(date %in% c(max(date), as.Date("2022-12-31"))) |>
  mutate(year = ifelse(year == min(year), "min_year", "max_year")) |>
  select(code, eng, year, production_yoy) |>
  pivot_wider(names_from = year, values_from = production_yoy) |>
  mutate(
    pct_10y_change = number(max_year / min_year, accuracy = 0.01),
    middle_year = exp((log(min_year) + log(max_year))/2),  # geomean
    color = ifelse(min_year > max_year, "#f95d6a", "#2f4b7c")
  ) |>
  filter(!grepl("^T", eng)) |>
  ggplot() +
  geom_segment(
    aes(x = min_year, xend = max_year, y = fct_reorder(eng, max_year), color = I(color)),
    linewidth = 1.2,
    lineend = 'round', linejoin = 'round',
    arrow = arrow(length = unit(0.1, "inches"))
  ) +
  geom_point(aes(x = min_year, y = eng), color = new_palette_colors[6], size = 3) +
  geom_point(aes(x = max_year, y = eng), color = new_palette_colors[2], size = 3) +
  geom_text(aes(x = middle_year, y = eng, label = pct_10y_change), vjust = 0) +
  geom_point(aes(x = 450, y = 2.5), color = new_palette_colors[6], size = 3) +
  geom_point(aes(x = 450, y = 1.5), color = new_palette_colors[2], size = 3) +
  geom_text(aes(x = 600, y = 2.5, label = 2013)) +
  geom_text(aes(x = 600, y = 1.5, label = 2023)) +
  scale_x_log10(labels = number_format()) +
  labs(
    x = "Billion AMD",
    y = NULL,
    title = "From 2013 to 2023, what area had the biggest growth in GDP?",
    subtitle = "The numbers show the growth rate",
    captions = paste0(caption_eng, "    |    Data Source: armstat.am")
  )

```



```{r, include=FALSE}

GDP_annual_pct <-
  GDP_annual |>
  filter(grepl("Domestic product", eng)) |>
  transmute(year, GDP_growth = vol_YoY_pct/100 - 1) |>
  na.omit()

GDP_annual_pct |>
  mutate(GDP_growth = GDP_growth + 1) |>
  summarise(GDP_mean_growth = (exp(mean(log(GDP_growth))) - 1)*100) |>
  pull(GDP_mean_growth)


GDP_annual_contribution <-
  GDP_annual |>
  left_join(GDP_annual_pct, by = "year") |>
  group_by(eng) |>
  mutate(
    vol_YoY_pct = vol_YoY_pct/100,
    contribution = lag(production) * (vol_YoY_pct - 1),
    contribution = ifelse(grepl("gross,", eng), NA, contribution)
  ) |>
  group_by(year) |>
  mutate(
    contribution = contribution / sum(contribution, na.rm = TRUE) * GDP_growth,
    contribution = ifelse(grepl("Domestic product", eng), vol_YoY_pct - 1, contribution)
  ) |>
  select(-GDP_growth)


GDP_contribution_plot_generator <-
  function(tbl, year_, groupped_codes_ = c("H","I","L","M","N","P","Q","S","T")){

    tbl_groupped <-
      tbl |>
      filter(year %in% year_) |>
      filter(!is.na(contribution)) |>
      mutate(
        groupped_codes = case_when(
          code %in% c("C", "D", "E") ~ "C+D+E",
          code %in% groupped_codes_ ~ "Servises\n(Other)",  # R ีฐีกีถีฅีฌ
          grepl("Taxes on products", eng) ~ "TA",
          grepl("Financial Intermediate", eng) ~ "FISIM",
          grepl("Domestic product", eng) ~ "GDP\nGrowth",
          TRUE ~ code
        ),
        groupped_eng = case_when(
          code %in% c("C", "D", "E") ~ "Other industries, including energy",
          code %in% groupped_codes_ ~ paste(groupped_codes_, collapse = "+"),
          grepl("Domestic product", eng) ~ NA,
          TRUE ~ eng
        ),
        groupped_eng = str_remove(groupped_eng, "; compulsory social security"),
        groupped_eng = str_remove(groupped_eng, " .FISIM."),
      ) |>
      group_by(groupped_codes, year, groupped_eng) |>
      summarise(
        contribution = sum(contribution),
        .groups = "drop"
      ) |>
      group_by(year) |>
      arrange(year) |>
      mutate(
        labels = percent(contribution, accuracy = 0.1),
        groupped_codes = fct_reorder(groupped_codes, contribution),
        groupped_codes = fct_relevel(
          groupped_codes, "TA", "FISIM", "GDP\nGrowth", after = Inf
        ),
        id = as.numeric(groupped_codes)
      ) |>
      arrange(id) |>
      mutate(
        annotation = paste(groupped_codes, groupped_eng, sep = " โ "),
        annotation = str_replace(annotation, "\n", " "),
        annotation = ifelse(grepl("GDP", annotation), " ", annotation),
        annotation = fct_reorder(annotation, id)
      ) |>
      mutate(
        end = cumsum(contribution),
        start = end - contribution,
        end = ifelse(groupped_codes == "GDP\nGrowth", 0, end),
        start = ifelse(groupped_codes == "GDP\nGrowth", contribution, start),
        fill_ = case_when(
          groupped_codes == "GDP\nGrowth" ~ "Total",
          contribution < 0 ~ "Negative",
          TRUE ~ "Positive"
        )
      )

    plot <-
      tbl_groupped |>
      ggplot() +
      geom_hline(yintercept = 0, color = new_palette_colors[3]) +
      geom_rect(
        aes(xmin = id - 0.45, xmax = id + 0.45,
            ymin = end, ymax = start, fill = fill_, linetype = annotation)
      ) +
      geom_text(aes(x = groupped_codes, y = (end+start)/2, label = labels)) +
      geom_segment(
        aes(x = id - 0.45, xend = ifelse(id == max(id), id + 0.45, id + 0.45 + 1),
            y = end, yend = end),
        linetype = 1, color = new_palette_colors[3]
      ) +
      scale_y_continuous(labels = percent_format(), n.breaks = 6) +
      scale_fill_manual(values = new_palette_colors[c(6,2,8)]) +
      labs(
        x = NULL,
        y = NULL,
        fill = NULL,
        linetype = NULL,
        subtitle = "Percentage Points",
        captions = paste0(caption_eng, "    |    Data Source: armstat.am"),
        # title = paste("Factors Contributing to", year_, "Economic Growth")
        title = paste("Forecast of Factors that will Contribute to", year_, "Economic Growth")
      ) +
      guides(
        linetype = guide_legend(override.aes = list(fill = "#FFFFFF00")),
        color = guide_legend(order = 2),
        fill = guide_legend(order = 1),
      ) +
      theme(
        panel.grid.major.x = element_blank(),
        legend.text = element_text(size = 9),
      )

    return(plot)

  }

GDP_contribution_plot_generator(GDP_annual_contribution, year_ = 2023)
```



```{r, include=FALSE}

biggest_gdp_contributers <-
  GDP_quarter |>
  mutate(
    date = yq(date) + months(3) - days(1),
    year = year(date)
  ) |>
  filter(
    !is.na(code),
    year == max(year)
  ) |>
  group_by(code) |>
  summarise(production = sum(production)) |>
  slice_max(order_by = production, n = 9) |>
  pull(code)

GDP_quarter |>
  # filter(!is.na(code) | grepl("Domestic product .gross", eng)) |>
  filter(code %in% biggest_gdp_contributers) |>
  # filter(grepl("Domestic product .gross", eng)) |>
  mutate(
    vol_YoY_pct = vol_YoY_pct / 100,
    date = yq(date) + months(3) - days(1),
    year = year(date),
    eng = str_trunc(eng, 40),
    eng = ifelse(is.na(code), eng, paste0(code, ". ", eng))
  ) |>
  group_by(code, eng) |>
  mutate(
    production_yoy = roll_sumr(production, 4) / 1e3,
    vol_pct_2 = production_yoy / lag(production_yoy, n = 4) - 1,
    vol_YoY_pct = vol_YoY_pct - 1
  ) |>
  ungroup() |>
  select(-code, -arm, -rus) |>
  ggplot(aes(date, vol_YoY_pct, color = eng)) +
  facet_wrap(~eng, scales = "free_y") +
  geom_hline(yintercept = 0, color = "gray") +
  geom_smooth(method = "glm") +
  geom_line() +
  scale_y_continuous(labels = percent_format()) +
  labs(
    x = NULL,
    y = NULL
  ) +
  theme(
    legend.position = "none"
  )

```



2023 ีฉีพีกีฏีกีถีซ 4ึึีค ีฅีผีกีดีฝีตีกีฏีซ ีฟีถีฟีฅีฝีกีฏีกีถ ีกีณีจ ีพีฅึีกีถีกีตีพีฅีฌ ีง 7.7 ีฟีธีฏีธีฝีซึ 6.4 ีฟีธีฏีธีฝีซึ ิฑีตีฝีซีถึีถ 1.3 ีฟีธีฏีธีฝีกีตีซีถ ีฏีฅีฟีธีพ ีฟีถีฟีฅีฝีกีฏีกีถ ีกีณีจ ีซีปีฅึีพีฅีฌ ีงึ ีีซีทีฅึีถีฅีถึ ีธึ ีกีตีฝ ีฎีกีพีกีฌีกีตีซีถ ีกีณีซ ึีธึึีกีถีซีทีจ ีธึีถีซ ีฟีกึีฅีฏีกีถ ีฏีฟึีพีกีฎึ ึ ีฐีกีทีพีกึีฏีพีธึีด ีง ีถีกีญีธึีค ีฟีกึีพีก ีฐีกีดีกีบีกีฟีกีฝีญีกีถ ีฅีผีกีดีฝีตีกีฏีซ ีถีฏีกีฟีดีกีดีขึ ีีถีฟีฅีฝีกีฏีกีถ ีกีณีซ ีดีฅีป ีฏึีณีกีฟีพีฅีฌ ีง ีฐีฅีถึ ีธีฝีฏีธึ ีพีฅึีกีกึีฟีกีฐีกีถีดีกีถ "ีกึีคีตีธึีถีกีขีฅึีธึีฉีตีธึีถีจ", ีธึีจ ีฉีกึ ีกีผีกีพ 2023 ีฉีพีกีฏีกีถีซ ีถีธีตีฅีดีขีฅึีซึ ึ ีฆีฃีกีฌีซ ีธึีชีธีพ ีทีกึีธึีถีกีฏีพีฅึ ีดีซีถีนึ 2024 ีฉีพีกีฏีกีถีซ ีดีกีตีซีฝึ ิฑีตีฝีซีฝีธีพ ีพีฅึีกีถีกีตีพีฅีฌ ีง ีฐีซีดีถีกีฏีกีถีธึีด ีกึีคีตีธึีถีกีขีฅึีธึีฉีตีกีถ (C) ีกีณีจ ึ 1.3 ีฟีธีฏีธีฝีกีตีซีถ ีฏีฅีฟีจ ีฐีฅีถึ ีฃีกีฌีซีฝ ีงึ ีกึีคีตีธึีถีกีขีฅึีธึีฉีตีธึีถีซึ, ีธึีจ ีฆึีธีตีกึีพีฅีฌ ีงึ 

ีีทีฅีถึ ีธึ 2023 ีฉีพีกีฏีกีถีซ ีฟีกึีฅีฏีกีถ ีฟีถีฟีฅีฝีกีฏีกีถ ีกีณีซ ึีธึึีกีถีซีทีจ ีคีฅีผึีฝ ีถีทีพีกีฎ ีง 8.7 ีดีกีฏีกึีคีกีฏีซ ีพึีก ึ ีคีฅีผึีฝ ีนีซ ีพีฅึีกีถีกีตีพีฅีฌึ ีีฅึ ีฏีธีฒีดีซึ 2023 ีฉีพีกีฏีกีถีซ ีพีฅึีปีซีถ (citation needed) ีฏีกีถีญีกีฟีฅีฝีพีธึีด ีงึ ีธึ ีฟีถีฟีฅีฝีกีฏีกีถ ีกีณีจ ีฏีกีฆีดีฅีฌีธึ ีง ีดีธีฟ 7 ีฟีธีฏีธีฝ, ีฝีกีฏีกีตีถ ีถีธีตีฅีดีขีฅึีซึ ีธีฝีฏีธึ ีพีฅึีกีกึีฟีกีฐีกีถีดีกีถ ีฆีฃีกีฌีซ ีดีฅีฎีกึีธึีดีจ ีธึีณีกึึีฅึ ีฟีถีฟีฅีฝีกีฏีกีถ ีกีณีจ 8.7 ีฟีธีฏีธีฝ ีจีถีคีฐีกีถีธึึ ีฟีถีฟีฅีฝีกีฏีกีถ ีกีถีฏีดีกีถ ึีธีถีซีถ ึ ีกีผีกีปีกึึีฅึ ีฉีกีถีฏีกึีชีฅึ ีงึึีฅีฏีฟึ ิฟีกึีฅีฌีซ ีง ีฝีบีกีฝีฅีฌ ีธึ 2023 ีฉีพีกีฏีกีถีซ 8.7 ีฟีธีฏีธีฝ ีฟีกึีฅีฏีกีถ ีฟีถีฟีฅีฝีกีฏีกีถ ีกีณีจ ีถีธึีตีถีบีฅีฝ ีฏีกึีธีฒ ีง ีพีฅึีกีถีกีตีพีฅีฌ ีคีฅีบีซ 7.5 ีฟีธีฏีธีฝึ


################################



ีีกีตีกีฝีฟีกีถีธึีด ีฅึีฏีกึีกีชีกีดีฏีฅีฟ ีดีซีปีซีถ ีฟีถีฟีฅีฝีกีฏีกีถ ีกีณีจ ีฏีกีฆีดีธึีด ีง 4.5 ีฟีธีฏีธีฝึ ีีกีถีฏีกึีกีฎ ีฐีกีดีกีฏีกึีฃ ีฑีฃีฟีธึีด ีง ีซึ ีฅึีฏีกึีกีชีกีดีฏีฅีฟ ีดีซีปีซีถีซีถ, ีฅีฉีฅ ีฐีกีดีกีฏีกึีฃีธึีด ีกีผีฏีก ีนีฅีถ ีธึีกีฏีกีฏีกีถ ึีธึีธีญีธึีฉีตีธึีถีถีฅึึ ีึีกีฏีกีฏีกีถ ึีธึีธีญีธึีฉีตีธึีถีถีฅึีจ ีถึีกีถึ ีฅีถ, ีธึีธีถึ ีฏีกีญีพีกีฎ ีนีฅีถ ีกึีฟีกึีซีถ ีฃีธึีฎีธีถีถีฅึีซ ีฐีฅีฟ, ีฏีกึีธีฒ ีฅีถ ีกีบีกีฐีธีพีพีฅีฌ ีกีพีฅีฌีกึีพีกีฎ ีกึีชีฅึีซ ีฏีกีตีธึีถ ีฅึีฏีกึีกีชีกีดีฏีฅีฟ ีฝีฟีฅีฒีฎีธึีดึ ิฑีตีฝีบีซีฝีซ ีธึีกีฏีกีฏีกีถ ึีธึีธีญีธึีฉีตีธึีถีถีฅึีจ ีฐีซีดีถีกีฏีกีถีธึีด ีฏีกีบีพีกีฎ ีฅีถ ีซีคีถีธึีฝีฟึีซีกีตีซ ึ ีซีถีคีธึีฝีฟึีกีฌีซีฆีกึีดีกีถ ีฐีฅีฟึ

ีีกีตีกีฝีฟีกีถีธึีด ีฟีถีฟีฅีฝีกีฏีกีถ ีกีณีจ ีฐีซีดีถีกีฏีกีถีธึีด ีกีบีบีกีฐีพีฅีฌ ีง ีกึีฟีกึีซีถ ีฃีธึีฎีธีถีถีฅึีธีพี 2022 ีฉีพีกีฏีกีถีซีถ ีีธึีฝีกีฟีกีถีซึ ีดีฅีฎ ีฎีกีพีกีฌีธีพ ีฏีกีบีซีฟีกีฌีซ ีถีฅึีฐีธีฝึีธีพ, ีซีฝีฏ 2023 ีฉีพีกีฏีกีถีซ ีถีธีตีฅีดีขีฅึีซึี ีดีฅีฎ ีฎีกีพีกีฌีถีฅึีธีพ ีธีฝีฏีธึ ีพีฅึีกีกึีฟีกีฐีกีถีดีกีดีขึ ีีถีฟีฅีฝีกีฏีกีถ ีกีณีจ ีฐีซีดีถีกีฏีกีถีธึีด ีกีบีกีฐีธีพีพีธึีด ีง ีกีผึีฟึีซ, ีกีถีทีกึีช ีฃีธึีตีงีซ ึ ึีซีถีกีถีฝีกีฏีกีถ ีฎีกีผีกีตีธึีฉีตีธึีถีถีฅึีซ ีฏีธีฒีดีซึ, ีธึีธีถึ ีฏีกีญีพีกีฎ ีฅีถ ีกึีฟีกึีซีถ ีฃีธึีฎีธีถีถีฅึีซ ีฐีฅีฟึ

ิฑีตีฝีบีซีฝีซ ีบีกีตีดีกีถีถีฅึีธึีด ีญีฅีฌีกีดีซีฟ ีนีง ีญีธีฝีฅีฌ ีีกีตีกีฝีฟีกีถีซ ีฏีกีผีกีพีกึีธึีฉีตีกีถ ีฏีธีฒีดีซึ ีฉีซึีกีญีกีพีธึีกีฎ 7%-ีซ ีฅึีฏีกึีกีชีกีดีฏีฅีฟ ีฟีถีฟีฅีฝีกีฏีกีถ ีกีณีซ ีดีกีฝีซีถึ ิปีถีนีบีฅีฝ ีฅึึีธึีด ีง 4-ึีค ีฃีฎีกีบีกีฟีฏีฅึีธึีด 2020 ีฉีพีกีฏีกีถีซ ิฟีีิปิด-ีซ ีฐีกีดีกีณีกึีกีฏีซ ึ 44-ึึีตีก ีบีกีฟีฅึีกีฆีดีซ ีคีฅีบึีฅึีซึ ีฐีฅีฟีธ ีฟีถีฟีฅีฝีกีฏีกีถ ีพีฅึีกีฏีกีถีฃีถีธึีดีซึ ีฐีฅีฟีธ ีดีฅึ ีฟีถีฟีฅีฝีธึีฉีตีกีถ ีฐีซีดีถีกีฏีกีถ ีณีตีธึีฒีฅึีจ ีกีบีกีฐีธีพีฅีฌ ีฅีถ ีฟีถีฟีฅีฝีกีฏีกีถ ีกีณีซ ีดีธีฟ 4 .5 ีฟีธีฏีธีฝีจ, ีธึีซึ ีดีธีฟ 1-2.5 ีฟีธีฏีธีฝีกีตีซีถ ีฏีฅีฟีจ ีกีบีกีฐีธีพีฅีฌ ีง ีี (J) ีธีฌีธึีฟ` ีฐีซีดีถีฏีกีถีกึีธีด ีซ ีฐีกีทีซีพ ีผีธึีฝีกีฏีกีถ ีี ีดีกีฝีถีกีฃีฅีฟีถีฅึีซ ีผีฅีฌีธีฏีกึีซีกีตีธีพึ ิฑึีคีฅีถ 3-ึีค ีฅีผีกีดีฝีตีกีฏีถ ีง ีกีตีฝ ีธีฌีธึีฟีจ ีกีถีฏีดีกีถ ีดีฅีป ีงึ ีีซีดีถีกีฏีกีถ ีกีณีจ ีกีบีกีฐีธีพีฅีฌ ีฅีถ ึีซีถีกีถีฝีฏีกีถ ีฎีกีผีกีตีธึีฉีตีธึีถีถีฅึีจ (K), ีกีตีฝีซีถึีถ ีขีกีถีฏีกีตีซีถ ีฐีกีดีกีฏีกึีฃีจ ึ ีกีผึีฟีธึึีจ, ีถีทีฅีถึ ีธึ ีีธึีฝีกีฟีกีถีซึ ีดีฅีฎ ีฎีกีพีกีฌีธีพ ีฏีกีบีซีฟีกีฌีซ ีถีฅึีฐีธีฝึีซ ีกึีคีตีธึีถึีธึีด ีขีกีถีฏีกีตีซีถ ีฐีกีดีกีฏีกึีฃีซ ีฆีธึีฟ ีทีกีฐีธึีตีฉีจ 2022 ีฉีพีกีฏีกีถีซีถ ีกีณีฅึ 3 ีกีถีฃีกีดึ ิปีถีนีบีฅีฝ ีฅึึีธึีด ีง 4-ึีค ีฃีฎีกีบีกีฟีฏีฅึีซึ ีฐีซีดีถีกีฏีกีถ ีธีฌีธึีฟีถีฅึีซ ีฏีกีตีธึีถ ีถีบีกีฝีฟีธึีดีจ ีฟีถีฟีฅีฝีกีฏีกีถ ีกีณีซีถ ีฝีฏีฝีพีฅึ 2022 ีฉีพีกีฏีกีถีซ ีฝีฏีฆีขีซึ ีดีซีถีนึ 2023 ีฉีพีกีฏีกีถีซ ีฏีฅีฝีจ, ีธึีซึ ีฐีฅีฟีธ ีฃึีฅีฉีฅ ีขีธีฌีธึ ีฟีถีฟีฅีฝีกีฏีกีถ ีฃีธึีฎีธึีถีฅีธึีฉีตีกีถ ีคีกีฝีกีฏีกึีฃีซีนีถีฅึีจ ีกีถีฏีธึีด ีฅีถ ีฃึีกีถึีธึีด ีจีถีคีฐีกีถีธึึ ีฟีถีฟีฅีฝีกีฏีกีถ ีกีณีซ ีถีบีกีฝีฟีดีกีถ ีดีฅีปึ 

ิปีถีนีบีฅีฝ ีฅึึีธึีด ีง 5-ึีค ีฃีฎีกีบีกีฟีฏีฅึีธึีด 2024 ีฉีพีกีฏีกีถีซ 2-ึีค ีฅีผีกีดีฝีตีกีฏีซ 6.4 ีฟีธีฏีธีฝ ีฟีถีฟีฅีฝีกีฏีถ ีกีณีซ ีฃึีฅีฉีฅ ีฏีฅีฝีจี 3.1 ีฟีธีฏีธีฝีกีตีซีถ ีฏีฅีฟีจ ีกีบีกีฐีธีพีธึีด ีง ีดีฅีฎีกีฎีกีญ ึ ีดีกีถึีกีฎีกีญ ีกีผึีฟีธึึีจ (G)ึ ีีธึีทีกีฃึีกีพีถ ีกีตีถ ีง ีธึ 2-ึีค ีฐีธึีซีฆีธีถีกีฏีกีถีธึีด ีง ีกีถีทีกึีช ีฃีธึีตึีซ ีฐีฅีฟ ีฏีกีบีพีกีฎ ีฃีธึีฎีกึึีถีฅึีจ, ีธึีจ ีกีบีกีฐีธีพีฅีฌ ีง ีฟีถีฟีฅีฝีกีฏีกีถ ีกีณีซ ีดีธีฟ ีฅึึีธึีคีจ ีดีกีฝีจี 2.3 ีฟีธีฏีธีฝีกีตีซีถ ีฏีฅีฟีจึ ีีทีฅีถึ ีธึ ีดีซีถีนึ ีถีทีพีกีฎ ีฅีผีกีดีฝีตีกีฏ ีกีถีทีกีนีช ีฃีธึีตึีซ ีฐีฅีฟ ีฏีกีบีพีกีฎ ีฃีธึีฎีกึึีถีฅึีจ ีฏีกีฆีดีธึีด ีงีซีถ ีฟีถีฟีฅีฝีกีฏีกีถ ีกีณีซ ีถีบีกีฝีฟีดีกีถ ีดีซีปีซีถีจ ีฏีธีฝ ีฟีธีฏีธีฝีจ (ีซีถึีบีฅีฝ ีฅึึีธึีด ีง 3-ึีค ึ 4-ึีค ีฃีฎีกีบีกีฟีฏีฅึีธึีด)ึ


ิฑีถีทีกึีช ีฃีธึีตึีซ ีฐีฅีฟ ีฏีกีบีพีกีฎ ีฃีธึีฎีกึึีถีฅึีซ ีกีณีจ ีฃีกีฌีซีฝ ีง NACE Rev. 2 ีดีฅีฉีธีคีกีขีกีถีธึีฉีตีธึีถีซึ: ิฑีตีถ ีถีฅึีกีผีธึีด ีง ีถีกึ "ีีฅึีกีฏีกีถีกีฟีซึีธีป ีฏีธีฒีดีซึ ีฆีขีกีฒีฅึึีกีฎ ีขีถีกีฏีกึีกีถีถีฅึีซ ีฐีกีดีกึ ีฐีกีทีพีกึีฏีพีกีฎ ีพีกึีฑีกีฏีกีฌีธึีฉีตีกีถ ีกึีชีฅึีจ, ีธึีจ ีฃีถีกีฐีกีฟีพีธึีด ีง ีกีตีถ ีขีกีถีซ, ีฉีฅ ีฝีฅึีกีฏีกีถีกีฟีฅึีฅึีจ ีธึึีกีถ ีพีกึีฑีกีพีณีกึ ีฏีพีณีกึีฅีซีถ ีฐีกีดีกึีชีฅึ ีขีถีกีฏีกึีกีถีกีตีซีถ ีฎีกีผีกีตีธึีฉีตีธึีถีถีฅึีซ ีฐีกีดีกึ:" ิฑีตีฝีซีถึีถ ีีกีตีกีฝีฟีกีถีธึีด ีขีธีฌีธึ ีขีถีกีฏีกึีกีถีถีฅึีซ ีบีธีฟีฅีถึีซีกีฌ ีพีกึีฑีกีพีณีกึีจ ีกีถีฏีกีญ ีถึีกีถีซึ ีฝีฅึีกีฏีกีถีกีฟีฅึีถ ีก ีกีตีค ีบีกีฐีซีถ ีขีถีกีฏีกึีกีถีซ ีดีฅีป ีฉีฅ ีฟีถีพีธึีจึ ีีกีฝีฟีกึีซ 6.4 ีฟีธีฏีธีฝ ีกีณีซ ีฅึึีธึีค ีดีกีฝีจ ีกีบีกีฐีธีพีพีฅีฌ ีง ีกีถีทีกีนีช ีฃีธึีตึีซ ีทีธึีฏีกีตีกีฏีกีถ ีกึีชีฅึีซ ีขีกึีฑึีกึีดีกีดีข, ีธึีจ sustanable ีนีงึ

ีีทีฅีถึ ีธึ ีกีตีฝ ีบีกีฐีซีถ ีกีผีฏีก ีง ีฐีกึีฏีกีตีซีถ ีฅีฏีกีดีธึีฟีถีฅึีซ ีฆีฃีกีฌีซ 9 ีฟีธีฏีธีฝ ีฉีฅึีฐีกีพีกึีกีฃึีดีกีถ, ีธึีจ ีบีกีฟีณีกีผีถีฅึีจ ึีถีถีกึีฏีฅีฌ ีฅีถึ ีดีฅึ [๐งฎโณ๐ฒ ีีกีตีกีฝีฟีกีถีจ ีฐีกึีฏีธึีด ีง ีชีกีดีกีถีกีฏีจโค ิฝีกีฒ ีฟีถีฟีฅีฝีกีฏีกีถ ีกีณีซ ีฐีฅีฟ] ีฐีธีคีพีกีฎีธีซีด, ีฝีก ีกึีคีฅีถ  ึีถีถีกึีฏีพีธึีด ีฅีถ ีขีตีธึีปีฅีฟีกีตีซีถ ีฎีกีญีฝีฅึีซ ีฏึีณีกีฟีธึีดีจึ  ีีกึ ีบีกีฌีกีถีถีฅึีธึีด ีฏีก ีดีธีฟ 150 ีดีฌึีค ีคึีกีดีซ ีบีกีฐีธีฝีฟีกีตีซีถ ึีธีถีคีจ ึีฃีฟีกีฃีธึีฎีฅีฌ ีฅีฏีกีดีธึีฟีถีฅึีซ ีฉีฅึีฐีกีพีกึีกีฃึีดีกีถ ีบีกีตีดีกีถีถีฅึีธึีด ีฎีกีญีฝีฅึีจ ีกีบีกีฐีธีพีฅีฌีธึึ ิฑีตีฝ ีบีกีตีดีกีถีถีฅึีธึีด ีดีฅีฎ ีง ีบีฅีฟีกีฏีกีถ ีบีกึีฟึีซ ีดีฅีฎีกึีดีกีถ ีฐีกีพีกีถีกีฏีกีถีธึีฉีตีธึีถีจึ

ีีทีฅีถึ ีถีกึ ีธึ ีฅีฉีฅ ีนีกีบีกีฐีธีพีพีซ ีขีตีธึีชีฅีซ ีดีกีฝีซีถ ึึีฅีถึีธีพ ีถีกีญีกีฟีฅีฝีพีกีฎ 7 ีฟีธีฏีธีฝ ีฟีถีฟีฅีฝีกีฏีกีถ ีกีณีจ ีกีตีฌ ีฅึีฏีกึีกีชีกีดีฏีฅีฟ ีดีซีปีซีถีจี 4.5 ีฟีธีฏีธีฝีจ, ีกีบีก ีฏีกีผีกีพีกึีธึีฉีตีธึีถีจ ีกีตีฌ ีฐีกีพีกีฝีกึ ีบีกีตีดีกีถีถีฅึีธึีด ีฝีฟีซีบีพีกีฎ ีฏีฌีซีถีซ ีบีกีฐีธีฝีฟีกีตีซีถ ึีธีถีคีซึ ีดีธีฟ ีฏีฅีฝีจ ีฐีกีฟีฏีกึีถีฅีฌ ีฟีกึีขีฅึีธึีฉีตีกีถ ีกีบีกีฐีธีพีพีดีกีถ ีพึีกึ 2.5 ึ ีบีกีฐีกีฝ ีฟีธีฏีธีฝ ีฟีถีฟีฅีฝีกีฏีกีถ ีกีณีซ ีบีกีตีดีกีถีถีฅึีธึีด ีฝีฟีซีบีกีพีฎ ีฏีฌีซีถีฅีถึ ีดีฅีฎีกึีถีฅีฌ ีบีฅีฟีกีฏีกีถ ีบีกึีฟึีจึ

ีีิฑ ีกีณีซ ีดีฅึ ีฅึีฏีกึีกีชีกีดีฏีฅีฟ ีฏีกีถีญีกีฟีฅีฝีธึีดีถีฅึีจ ีกีตีฝ ีบีกีฐีซีถ ีคึีกีฏีกีถ ีนีฅีถึ ิฑีผีฏีก ีนีง ีฏีกีตีธึีถ ีฅึีฏีกึีกีชีกีดีฏีฅีฟ ีกีพีฅีฌีกึีพีกีฎ ีกึีชีฅึ ีกีบีกีฐีธีพีธีฒ ีกีณีซ ีฐีซีดึึ ิฑีตีฝ ีบีกีตีดีกีถีถีฅึีธึีด ีดีฅีฎ ีง ีบีฅีฟีกีฏีกีถ ีบีกึึีซ ีกีพีฅีฌีกึีดีกีถ ีพีฟีกีถีฃีจึ
